Cache is King: Presenting \corrfunc -- a Suite of Fast Correlation Function Codes

### 2016-07-20 paper/ms.tex
Working on the complexity

\usepackage{amsmath}
\usepackage{aas_macros}
\usepackage{rotating}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{color}
\usepackage{listings}
\usepackage{caption}
\usepackage{url}
\usepackage{adjustbox}
\journal{Astronomy and Computing}

%%% Definitions added by Manodeep 
\newcommand{\Msun}{M_{\odot}}
\newcommand{\hMsun}{h^{-1}M_{\odot}}
\newcommand{\sdss}{{\small{SDSS}}\xspace}
\newcommand{\hMpc}{\ensuremath{{h^{-1}Mpc}\xspace}}
\newcommand{\hkpc}{\ensuremath{{h^{-1}kpc}\xspace}}
\newcommand{\emcee}{\texttt emcee}
\newcommand{\rmax}{\ensuremath{{{\mathcal{R}}_{max}}}\xspace}
\newcommand{\rmaxcubed}{\ensuremath{{\mathcal{R}^{3}_{max}}}\xspace}
\newcommand{\xir}{\ensuremath{{DD(r)}}\xspace}
\newcommand{\xiofr}{\ensuremath{{\xi(r)}}\xspace}
\newcommand{\wprp}{\ensuremath{{w_p(r_p)}}\xspace}
\newcommand{\xirppi}{\ensuremath{{\xi(r_p,\pi)}}\xspace}
\newcommand{\cis}{\ensuremath{{pN(r)}}\xspace}
\newcommand{\ddrppi}{\ensuremath{{DD(r_p,\pi)}}\xspace}
\newcommand{\wtheta}{\ensuremath{{DD(\theta)}}\xspace}
\newcommand{\todo}[1]{\marginpar{TODO}{\color{red}#1}}
\newcommand{\avx}{\texttt AVX}
\newcommand{\sse}{\texttt SSE}

@@

\title{Cache is King: Presenting {\tt Corrfunc} -- a Suite of Fast Correlation Function Codes}

@@

\begin{abstract}
In the era of `Big Data', we need special computing techniques to both process the raw data as well 
generate and evaluate models to test our theories. In particular, large galaxy surveys
like the Sloan Digitial Sky Survey requires computing a variety of $n$-point statistics, e.g., the 
2-pt projected correlation function, $\wprp$. 
Since the clustering measure on the data is only done once, time taken to do
such a measurement is not a bottleneck. However, modeling the observed galaxy distribution correctly
{\em requires} a Monte-Carlo Markov Chain (MCMC) and repeated measurements of one or multiple
clustering statistics. In order to efficiently compute {\em any} clustering statistic,
we need to write code that is tuned for the underlying hardware. For
modern CPUs, such tuning involves proper utilization of the cache hierarchy {\em and}
vectorization with Single Instruction Multiple Data (SIMD) capable wide vector registers. 
In this paper, I present {\tt Corrfunc} --  a suite of OpenMP parallelized, clustering
codes that exploit current CPU micro-architecture with hand-written Advanced
Vector Extensions (\avx) and  Streaming SIMD Extensions (\sse) intrinsics. 
Given a 3-D distribution of points in a cartesian system (i.e.,
generated from cosmological simulations), {\tt Corrfunc} can compute spatial
3-D correlation functions \xir, \xiofr, \xirppi; 2-point 
projected correlation function \wprp; counts-in-spheres \cis. For a 3-D
particle distribution on the sky (i.e., mock galaxy catalogs, observed galaxies), the codes can compute 
projected correlation function \ddrppi, angular correlation function \wtheta as
well as \cis.  By design, {\tt Corrfunc} is highly optimized 
and can compute \wprp for $\mathcal{O}$(1 million) galaxies in
$\sim 6$ seconds on a post-2011 CPU, which is a factor of few faster than 
existing public correlation function routines.  {\tt Corrfunc} is
publicly available at \url{https://github.com/manodeep/Corrfunc/} while
benchmarks with existing codes are available at
\url{https://gist.github.com/manodeep/cffd9a5d77510e43ccf0}.
\end{abstract}

@@

\section{PACKAGE SUMMARY}
{\tt Corrfunc} is written primarily in {\tt C} and comes with convenient
{\tt python} wrappers for typical clustering statistics. The fundamental design
principles for {\tt Corrfunc} are the following:
\begin{enumerate}
  \item Correct -- {\tt Corrfunc} has a base set of correct outputs for every statistic
    generated either through slow, brute-force methods or independent, 
    external codes. Within {\tt Corrfunc}, every clustering statistic is covered with at least one
    test case that {\em requires} reproducing this `known-correct' result {\em
      exactly}. The `PASS/FAIL' status for each test is the direct output of {\tt diff -q
      test\_output known\_correct\_output}.  
  \item High Performance -- Performance is the overarching goal for {\tt
    Corrfunc}. The code-base has been repeatedly analyzed with Intel VTune
    (\url{https://software.intel.com/en-us/intel-vtune-amplifier-xe})
    on a variety of platforms to optimize the innermost loops. 
  \item Portable -- {\tt Corrfunc} is written in {\tt ISO/IEC 9899:1999}
    compliant {\tt C}. All instruction set specific codes are protected via compile-time constant
    definitions.
\end{enumerate}
Every clustering statistic in {\tt Corrfunc} can either be accessed through the
API call (e.g., via {\tt python}), linking with the static libraries) or as
a command-line executable.

\section{Methods}
\subsection{Complexity of the Code}
We need to compute pairwise~\citep{} distances to get the correlation function. A naive implementation of a correlation function would compute {\em all possible} 
pairwise separations with a complexity $\mathcal{O}(N^2)$. However, for almost all correlation functions, we are only interested in separations less than 
a certain \rmax, where \rmax is much smaller than the domain of the point distribution itself. We can then immediately see a way to prune pairs that can not
{\em possibly} be within \rmax. If we impose a 3-d grid, with cell-size \rmax, then two points separated by more than one cell size (\rmax) in any one 
dimension can not be within \rmax of each other. Thus, given one point which is the target galaxy and a grid with cell-size \rmax, 
immediately allows us to prune {\em all} of the points that are not within 1 cell offset along each dimension. However, even with this pruning, the actual 
implementation of the algorithm matters. For instance, a linked-list in each cell performs $\sim 60\times$ worse than the algorithm described here.\footnote{The 
usual linked-list is very cache-unfriendly. Each dereference requires a read
from a new region of memory and an almost guaranteed cache miss. } 
We can now compute the theoretical complexity of the proposed algorithm. Let
$\mathcal{L}$ be the side-length of the cube over which $\mathcal{N}$ points are distributed, then
each cell contains $r_{\mathrm max}^3\times \mathcal{N/V}$ particles. To compute the
correlation function, we have to loop over each point, and then {\em all} of
the points in the neighboring cells. This results in a complexity of 
$\mathcal{O(N{\dfrac{\rmax}/L}^3}$. A typical \rmax is $\sim 10-20\% \times
\mathcal{L}$, therefore ordering the particles in cells of size \rmax should
result in a speedup of $(\rmax/\mathcal{L})^3 \sim 125-1000$ compared to the
brute-force algorithm.

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{aas_macros}
\usepackage{rotating}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{color}
\usepackage{listings}
\usepackage{caption}
\usepackage{url}
\usepackage{adjustbox}
\journal{Astronomy and Computing}

%%% Definitions added by Manodeep 
\newcommand{\Msun}{M_{\odot}}
\newcommand{\hMsun}{h^{-1}M_{\odot}}
\newcommand{\sdss}{{\small{SDSS}}\xspace}
\newcommand{\hMpc}{\ensuremath{{h^{-1}Mpc}\xspace}}
\newcommand{\hkpc}{\ensuremath{{h^{-1}kpc}\xspace}}
\newcommand{\emcee}{\texttt emcee}
\newcommand{\corrfunc}{{\tt Corrfunc}\xspace}
\newcommand{\rmax}{\ensuremath{{{\mathcal{R}}_{max}}}\xspace}
\newcommand{\rmaxcubed}{\ensuremath{{\mathcal{R}^{3}_{max}}}\xspace}
\newcommand{\lbox}{\ensuremath{\mathcal{L}}\xspace}
\newcommand{\npart}{\ensuremath{\mathcal{N}}\xspace}
\newcommand{\xir}{\ensuremath{{DD(r)}}\xspace}
\newcommand{\xiofr}{\ensuremath{{\xi(r)}}\xspace}
\newcommand{\wprp}{\ensuremath{{w_p(r_p)}}\xspace}
\newcommand{\xirppi}{\ensuremath{{\xi(r_p,\pi)}}\xspace}
\newcommand{\cis}{\ensuremath{{pN(r)}}\xspace}
\newcommand{\ddrppi}{\ensuremath{{DD(r_p,\pi)}}\xspace}
\newcommand{\wtheta}{\ensuremath{{DD(\theta)}}\xspace}
\newcommand{\todo}[1]{\marginpar{TODO}{\color{red}#1}}
\newcommand{\avx}{\texttt AVX}
\newcommand{\sse}{\texttt SSE}

@@

\title{Cache is King: Presenting \corrfunc -- a Suite of Fast Correlation Function Codes}

@@

\begin{abstract}
In the era of `Big Data', we need special computing techniques to both process the raw data as well 
generate and evaluate models to test our theories. In particular, large galaxy surveys
like the Sloan Digitial Sky Survey requires computing a variety of $n$-point statistics, e.g., the 
2-pt projected correlation function, $\wprp$. 
Since the clustering measure on the data is only done once, time taken to do
such a measurement is not a bottleneck. However, modeling the observed galaxy distribution correctly
{\em requires} a Monte-Carlo Markov Chain (MCMC) and repeated measurements of one or multiple
clustering statistics. In order to efficiently compute {\em any} clustering statistic,
we need to write code that is tuned for the underlying hardware. For
modern CPUs, such tuning involves proper utilization of the cache hierarchy {\em and}
vectorization with Single Instruction Multiple Data (SIMD) capable wide vector registers. 
In this paper, I present \corrfunc --  a suite of OpenMP parallelized, clustering
codes that exploit current CPU micro-architecture with hand-written Advanced
Vector Extensions (\avx) and  Streaming SIMD Extensions (\sse) intrinsics. 
Given a 3-D distribution of points in a cartesian system (i.e.,
generated from cosmological simulations), \corrfunc can compute spatial
3-D correlation functions \xir, \xiofr, \xirppi; 2-point 
projected correlation function \wprp; counts-in-spheres \cis. For a 3-D
particle distribution on the sky (i.e., mock galaxy catalogs, observed galaxies), the codes can compute 
projected correlation function \ddrppi, angular correlation function \wtheta as
well as \cis.  By design, \corrfunc is highly optimized 
and can compute \wprp for $\mathcal{O}$(1 million) galaxies in
$\sim 6$ seconds on a post-2011 CPU, which is a factor of few faster than 
existing public correlation function routines.  \corrfunc is
publicly available at \url{https://github.com/manodeep/Corrfunc/} while
benchmarks with existing codes are available at
\url{https://gist.github.com/manodeep/cffd9a5d77510e43ccf0}.
\end{abstract}

@@

\section{PACKAGE SUMMARY}
\corrfunc is written primarily in {\tt C} and comes with convenient
{\tt python} wrappers for typical clustering statistics. The fundamental design
principles for \corrfunc are the following:
\begin{enumerate}
  \item Correct -- \corrfunc has a base set of correct outputs for every statistic
    generated either through slow, brute-force methods or independent, 
    external codes. Within \corrfunc, every clustering statistic is covered with at least one
    test case that {\em requires} reproducing this `known-correct' result {\em
      exactly}. The `PASS/FAIL' status for each test is the direct output of {\tt diff -q
      test\_output known\_correct\_output}.  
  \item High Performance -- Performance is the overarching goal for {\tt
    Corrfunc}. The code-base has been repeatedly analyzed with Intel VTune
    (\url{https://software.intel.com/en-us/intel-vtune-amplifier-xe})
    on a variety of platforms to optimize the innermost loops. 
  \item Portable -- \corrfunc is written in {\tt ISO/IEC 9899:1999}
    compliant {\tt C}. All instruction set specific codes are protected via compile-time constant
    definitions.
\end{enumerate}
Every clustering statistic in \corrfunc can either be accessed through the
API call (e.g., via {\tt python}), linking with the static libraries) or as
a command-line executable.

\section{Methods}
\subsection{Complexity of the Code}
We need to compute pairwise~\citep{} distances to get the correlation function. A naive implementation of a correlation function would compute {\em all possible} 
pairwise separations with a complexity $\mathcal{O}(N^2)$. However, for almost all correlation functions, we are only interested in separations less than 
a certain \rmax, where \rmax is much smaller than the domain of the point distribution itself. We can then immediately see a way to prune pairs that can not
{\em possibly} be within \rmax. If we impose a 3-d grid, with cell-size \rmax, then two points separated by more than one cell size (\rmax) in any one 
dimension can not be within \rmax of each other. Thus, given one point which is the target galaxy and a grid with cell-size \rmax, 
immediately allows us to prune {\em all} of the points that are not within 1 cell offset along each dimension. However, even with this pruning, the actual 
implementation of the algorithm matters. For instance, a linked-list in each cell performs $\sim 60\times$ worse than the algorithm described here.\footnote{The 
usual linked-list is very cache-unfriendly. Each dereference requires a read
from a new region of memory and an almost guaranteed cache miss. } 
We can now compute the theoretical complexity of the proposed algorithm. Let
\lbox be the side-length of the cube over which $\mathcal{N}$ points are distributed, then
each cell contains $r_{\mathrm max}^3\times \mathcal{N/V}$ particles. To compute the
correlation function, we have to loop over each point, and then {\em all} of
the points in the neighboring cells. This results in a complexity of 
$\mathcal{O(NM)}$, where $\mathcal{M} = \mathcal{N}\times\left(\rmax/\lbox\right)^3$. Typical \rmax is $\sim 0.10-0.2\times
\lbox$, therefore ordering the particles in cells of size \rmax should
result in a speedup of $(\rmax/\lbox)^3 \sim 125-1000$ compared to the
brute-force algorithm. Since every possible pair within \rmax has to be
examined, the algorithm deteriorates to $\mathcal{O(N}^2)$ for \rmax comparable
to \lbox. Space partitioning tree based approaches are far more suitable for
cases where \rmax is $\gtrsim \lbox/3$~\citep[e.g.,][]{mlpack2013,
  feng_kdcount}.
